### 2023-02-12 chapters/abstract.tex
introduction rewritten

	In interpretable machine learning, rule-based classifiers are particularly effective in representing the decision boundary using a set of rules. The interpretability of rule-based classifiers is generally related to the size of the rules, where smaller rules with higher accuracy are preferable in practice. As such, interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets. To this end, we propose an incremental learning framework, called {\imli}, which extends the classification problem to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning. \blue{Furthermore, rule-based classifiers suffer from limited expressiveness due to being confined to propositional logic. To learn more expressible yet interpretable classification rules, we propose a relaxation of rule-based classifiers based on logical formulas and  an efficient learning framework, called {\crr},  building on incremental learning and MILP (mixed integer linear programming). {\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.}

	\blue{Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning classifiers. In the presence of multiple fairness metrics for quantifying bias, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of formally verifying the bias of a classifier given the probability distribution of features. Additionally, we extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness. Fairness metrics globally quantify bias, but do not detect or explain its sources. To explain group-based fairness metrics, we propose fairness influence functions (FIF) with an aim of quantifying the influence of individual features and the intersection of multiple features  on the bias of a classifier. FIF  explains fairness by revealing potential individual or intersectional features attributing highly to the bias. Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.}

>>>>

	In interpretable machine learning, rule-based classifiers are particularly effective in representing the decision boundary using a set of rules. The interpretability of rule-based classifiers is generally related to the size of the rules, where smaller rules with higher accuracy are preferable in practice. As such, interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets. To this end, we propose an incremental learning framework, called {\imli}, which extends the classification problem to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning. Furthermore, rule-based classifiers suffer from limited expressiveness due to being confined to propositional logic. To learn more expressible yet interpretable classification rules, we propose a relaxation of rule-based classifiers based on logical formulas and  an efficient learning framework, called {\crr},  building on incremental learning and MILP (mixed integer linear programming). {\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.

	Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning classifiers. In the presence of multiple fairness metrics for quantifying bias, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of formally verifying the bias of a classifier given the probability distribution of features. Additionally, we extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness. Fairness metrics globally quantify bias, but do not detect or explain its sources. To explain group-based fairness metrics, we propose fairness influence functions (FIF) with an aim of quantifying the influence of individual features and the intersection of multiple features  on the bias of a classifier. FIF  explains fairness by revealing potential individual or intersectional features attributing highly to the bias. Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.

### 2023-02-12 chapters/fairness/fif/introduction.tex
introduction rewritten

Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias. Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}. In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset. To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier. Thus, \textit{we aim to design a bias attribution method that is global by construction}. In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features. Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature\cite{buolamwini2018gender,wang2022towards}. 
%\textit{By attributing bias to both individual and intersecting features, we can gain a more detailed understanding of bias} and pave the way for designing more effective fairness algorithms. 
In this paper, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.}

>>>>

Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias. Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}. In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset. To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier. Thus, \textit{we aim to design a bias attribution method that is global by construction}. In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features. Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}. 
%\textit{By attributing bias to both individual and intersecting features, we can gain a more detailed understanding of bias} and pave the way for designing more effective fairness algorithms. 
In this paper, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.}

### 2023-02-12 chapters/introduction.tex
introduction rewritten

\subsection*{Expressiveness via Logical Relaxation}
We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness~\cite{GMM2020}. Elaborately, we consider relaxed definitions of standard OR/AND operators in propositional logic by allowing exceptions in the construction of a clause and also in the selection of clauses in a rule. Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain and the higher expressiveness of Boolean cardinality constraints. Relaxed-CNF generalizes widely employed rule representations including CNF, DNF, and decision sets. While the combinatorial structure of relaxed-CNF rules offers exponential succinctness, the na\"ive learning techniques are computationally expensive. To this end, we propose an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules. Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.\todo{Up to this.}

\section{Fairness in Machine Learning}
As a technology machine learning is oblivious to societal good or bad. The success of machine learning as an accurate predictor, however, finds applications in high-stake decision-making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc. In such applications, the deployed  classifier often demonstrates bias towards certain \emph{sensitive demographic groups} involved in the data~\cite{dwork2012fairness}. For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}. Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier. For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}. To this end, different fairness enhancing algorithms have been devised to improve fairness with respect to one or multiple fairness metrics. These algorithms try to rectify and mitigate bias in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}. Researchers also study fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}. In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) identification of sources of unfairness.

@@

In order to solve the fairness verification problem, existing works have proposed two principled approaches.	Firstly,~\cite{albarghouthi2017fairsquare} propose a formal method approach to reduce the verification problem into the weighted volume computation of an SMT formula. Secondly,~\cite{bastani2019probabilistic} propose a sampling approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost. Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t. sensitive features and ignore correlations among non-sensitive features. As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remain major challenges.

In this thesis, we propose an efficient fairness verification framework for two classes of machine learning classifiers, classifiers represented as Boolean formulas and linear classifiers. Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas. {\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}. Because SSAT encoding allows separate quantification to each sensitive feature without any restriction on the number of features. In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic}, and more robust than the sample-based empirical verifiers~\cite{aif360-oct-2018}. 
%We also prove a finite-sample error bound on estimated fairness metrics, which is stronger than the existing asymptotic guarantees.

Linear classifiers have attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. Existing fairness verifier suffers from two-fold limitations for verifying linear classifiers: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations. Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification. {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming. To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network to represent the conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG). Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms. We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing fairness influences of a subset of features on shifting the incurred bias of the classifiers from the original bias.
%, which we study in detail in the following contribution of the thesis.

\subsection*{Identification of Sources of Unfairness}
While fairness metrics globally quantify bias, they cannot detect or explain the sources of bias~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}. In order to identify the sources of bias and also the effect of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which factors contribute how much to the bias of a classifier on a dataset}. To this end, we follow a feature-attribution approach to understand the sources of bias~\cite{begley2020explainability,lundberg2020explaining}, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. Particularly, we define and compute \textit{Fairness Influence Function} (FIF) that quantifies the contribution of individual and subset of features to the resulting bias. FIFs do not only allow practitioners to identify the features to act up on but also to quantify the effect of various affirmative~\cite{calmon2017optimized,hardt2016equality,kamiran2012decision,zemel2013learning,zhang2018mitigating,zhang2018fairness,zhang2019faht} or punitive actions~\cite{hua2021human,mehrabi2020exacerbating,solans2020poisoning} on the resulting bias. We also instantiate an algorithm, {\fairXplainer}, that uses variance decomposition among the subset of features and a local regressor to compute FIFs accurately, while also capturing the intersectional effects of the features. Our experimental analysis validates that {\fairXplainer} captures the influences of both individual features and higher-order feature interactions, estimates the bias more accurately than existing local explanation methods, and detects the increase/decrease in bias due to affirmative/punitive actions in the classifier.

\section{Tools}
	We develop following open-source tools for interpretable and fair machine learning.
	\begin{itemize}
		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for interpretable machine learning
		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for fairness verification and FIF computation
	\end{itemize}
\section{Outline}
We organize the thesis in the following way. We discuss preliminaries in Chapter~\ref{chapter:preliminaries}. In the first part of the thesis, we discuss an incremental learning framework based on MaxSAT solving for learning interpretable classifiers. We conclude this part by applying incremental learning on a more expressible but interpretable rule-based classifier in Chapter~\ref{chapter:crr}. In the second part of this thesis, we discuss  fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}. We conclude this part by discussing the identification of sources of unfairness of machine learning classifiers in Chapter~\ref{chapter:fif}. Finally, we conclude our thesis in Chapter~\ref{chapter:conclusion}.

>>>>

\subsection*{Expressiveness via Logical Relaxation}
We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness~\cite{GMM2020}. Elaborately, we consider relaxed definitions of standard OR/AND operators in propositional logic by allowing exceptions in the construction of a clause and also in the selection of clauses in a rule. Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain and the higher expressiveness of Boolean cardinality constraints. Relaxed-CNF generalizes widely employed rule representations including CNF, DNF, and decision sets. While the combinatorial structure of relaxed-CNF rules offers exponential succinctness, the na\"ive learning techniques are computationally expensive. To this end, we propose an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules. Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.

\section{Fairness in Machine Learning}
As a technology machine learning is oblivious to societal good or bad. The success of machine learning as an accurate predictor, however, finds applications in high-stake decision-making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc. In such applications, the deployed  classifier often demonstrates bias towards certain \emph{sensitive demographic groups} involved in the data~\cite{dwork2012fairness}. For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}. Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier. For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more frequently than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}. To this end, different fairness enhancing algorithms have been devised to improve fairness with respect to one or multiple fairness metrics. These algorithms try to rectify and mitigate bias in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}. Researchers also study fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}. In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) interpreting or explaining of sources of unfairness.

@@

In order to solve the fairness verification problem, existing works have proposed two principled approaches.	Firstly,~\cite{albarghouthi2017fairsquare} propose a formal method approach to reduce the verification problem into the weighted volume computation of an SMT formula. Secondly,~\cite{bastani2019probabilistic} propose a sampling approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost. Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t.\ sensitive features and ignore correlations among non-sensitive features. As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remain major challenges.

In this thesis, we propose an efficient fairness verification framework for two classes of machine learning classifiers, classifiers represented as Boolean formulas and linear classifiers. Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas. {\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}. Because SSAT encoding, by construction, allows separate quantification to each sensitive feature without any restriction on the number of features, thereby extending the verification problem to multiple sensitive features unlike earlier methods. In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic} because of the efficient SSAT encoding.
%, and more robust than the sample-based empirical verifiers~\cite{aif360-oct-2018}. 
%We also prove a finite-sample error bound on estimated fairness metrics, which is stronger than the existing asymptotic guarantees.

Linear classifiers have attracted significant attention from researchers in the context of designing and testing prototype fairness enhancing and attack algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. In the context of verifying the bias of linear classifiers, existing fairness verifiers suffer from two-fold limitations: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations. To alleviate these limitations, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification. {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming. To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network to represent the conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG). Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics. We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing fairness influences of a subset of features on shifting the incurred bias of the classifiers from the original bias.
%, which we study in detail in the following contribution of the thesis.

\subsection*{Explaining Fairness metrics: Identifying the Sources of Bias}
Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias. Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}. In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset. To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier. Thus, \textit{we aim to design a bias attribution method that is global by construction}. In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features. Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}. In this thesis, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.} Building on global variance decomposition, we propose an algorithm {\fairXplainer} to estimate the fairness influence function (FIF) of features. In experiments, {\fairXplainer} approximates bias based on FIFs more accurately than existing methods. We also demonstrate higher correlation of FIFs with fairness interventions, thereby demonstrating the importance of FIFs in designing improved fairness algorithms in future.

\section{Tools}
	We develop following open-source tools for interpretable and fair machine learning.
	\begin{itemize}
		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for learning interpretable rule-based classification rules in a supervised learning setting.
		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
		\item \red{FairXplainer for explaining group-based fairness metrics in machine learning.}
	\end{itemize}
\section{Outline}
We organize the thesis in the following way. We discuss preliminaries in Chapter~\ref{chapter:preliminaries}. In the first part of the thesis on interpretable machine learning, we discuss an incremental learning framework based on MaxSAT for learning interpretable rule-based classifiers. We conclude this part by applying incremental learning on a more expressible yet interpretable rule-based classifier in Chapter~\ref{chapter:crr}. In the second part of the thesis on fairness in machine learning, we discuss probabilistic fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}. We conclude this part by explaining group fairness metrics in Chapter~\ref{chapter:fif}. Finally, we conclude our thesis in Chapter~\ref{chapter:conclusion}.
